---
title: "United States Historical Weather Events and their Impacts"
author: "Ahsan Habib Sunny"
date: "`r format(Sys.Date(),'%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Synopsis

Purpose of this analysis is to understand the actual impact of historical weather events in United States between 1950 and November of 2011. For this, we use the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. The analysis determine the impact in population health (considering injuries and fatalities) by EVTYPE(Event Type). Besides this, the analysis measures impact of these weather events over Economic across the US, considering damage in properties and crops during the years. At the very end of the analysis, we can predict that the weather event with more fatalities and injuries is the tornado, but overall the most important event in terms of the cost is the flood.

# Data Processing


1. Loading Data

#load packages

```{r}

library(dplyr)

library(gridExtra)

library(ggplot2)

```


```{r}

setwd("C:\\Users\\A S U S\\Desktop\\project")

stormdata<- read.csv("repdata_data_StormData.csv",header=T)

```

2. Processing data considering health and economic consequences

The only columns we need for the analysis are the ones related to the weather events. And their damages: “EVTYPE”, “FATALITIES”, “INJURIES”, “PROPDMG”, “PROPDMGEXP”,“CROPDMG” and “CROPDMGEXP”.


```{r}

newdata<- stormdata[,c(8,23:28)]

# Subset the data set for "FATALITIES" and "INJURIES" analysis

FTS <- select(newdata,c(EVTYPE,FATALITIES))

INS <- select(newdata,c(EVTYPE,INJURIES))
```

3. Processing the data considering economic consequences

```{r}

PRP <- arrange(count(select(newdata,c(EVTYPE,PROPDMG,PROPDMGEXP)),PROPDMGEXP),desc(n))
PRP <- slice(PRP,-1)

CRP<- arrange(count(select(stormdata,c(EVTYPE,CROPDMG,CROPDMGEXP)),CROPDMGEXP),desc(n))
CRP<- slice(CRP,-1)

# Calculate the percentage of the different cases of PROPDMGEXP and CROPDMGEXP

PRP$Percentage <- as.numeric(round((PRP$n/sum(PRP$n))*100,2))
SumPRP <- sum(PRP[PRP$PROPDMGEXP=="K"|PRP$PROPDMGEXP=="B"|PRP$PROPDMGEXP=="M",]$Percentage)

CRP$Percentage<- as.numeric(round((CRP$n/sum(CRP$n))*100,2))
SumCRP<- sum(CRP[CRP$CROPDMGEXP=="K"|CRP$CROPDMGEXP=="B"|CRP$CROPDMGEXP=="M",]$Percentage)
```

Therefore, the PROPDMG cases labeled by “K”, “M”, or “B” are 99.92483% of the cases and for CROPDMG are 99.98274% of the cases. As a result, I'm going to use these labels.


```{r}
# Subset the data set for "PROPDMG" and "CROPDMG" analysis

nd1<- newdata %>% select(c(EVTYPE,PROPDMG,PROPDMGEXP)) %>% filter(PROPDMGEXP=="K"|PROPDMGEXP=="M"|PROPDMGEXP=="B")
nd1$Cost <- ifelse(nd1$PROPDMGEXP=="K",nd1$PROPDMG*1000,ifelse(nd1$PROPDMGEXP=="M",nd1$PROPDMG*1000000,nd1$PROPDMG*1000000000))

nd2<- newdata %>% select(c(EVTYPE,CROPDMG,CROPDMGEXP)) %>% filter(CROPDMGEXP=="K"|CROPDMGEXP=="M"|CROPDMGEXP=="B")
nd2$Cost <- ifelse(nd2$CROPDMGEXP=="K",nd2$CROPDMG*1000,ifelse(nd2$CROPDMGEXP=="M",nd2$CROPDMG*1000000,nd2$CROPDMG*1000000000))
```

# Results
1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


```{r}
# Population Health consequences over EVTYPE

#Considering Fatalities 

RFTS <- FTS %>% group_by(EVTYPE) %>% summarise(Fatalities= sum(FATALITIES)) %>% arrange(desc(Fatalities))
RFTS <- slice(RFTS,1:25)


# Plot Fatalities

RFTS$EVTYPE <- factor(RFTS$EVTYPE,levels=RFTS$EVTYPE[order(-RFTS$Fatalities)])

gRFTS <- ggplot(RFTS,aes(EVTYPE,Fatalities),fill=Fatalities)+geom_bar(stat="identity",fill="blue",color="black")+ theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+labs(title="Total Number of Fatalities by Event Type")+xlab("Event Type")+ylab("Number of Fatalities")+theme(plot.title = element_text(hjust = 0.5,size = 10, face = "bold"))

#Considering INJURIES

RINS <- INS %>% group_by(EVTYPE) %>% summarise(Injuries= sum(INJURIES)) %>% arrange(desc(Injuries))
RINS <- slice(RINS,1:25)

#Plot Injuries

RINS$EVTYPE <- factor(RINS$EVTYPE,levels=RINS$EVTYPE[order(-RINS$Injuries)])

gRINS <- ggplot(RINS,aes(EVTYPE,Injuries),fill=Injuries)+geom_bar(stat="identity",fill="blue",color="black")+ theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+labs(title="Total Number of Injuries by Event Type")+xlab("Event Type")+ylab("Number of Injuries")+theme(plot.title = element_text(hjust = 0.5,size = 10, face = "bold"))

# Plot graph together

grid.arrange(gRFTS,gRINS,ncol=2)

```

2. Across the United States, which types of events have the greatest economic consequences?


```{r}
# Considering PROPDMG

nd3<- nd1 %>% group_by(EVTYPE) %>% summarise(Cost= sum(Cost)) %>% arrange(desc(Cost))

nd4<- slice(nd3,1:25)

# Transfor the cost column into "Billion dollars"
 
nd4$Cost <- nd4$Cost/1000000000

nd4$EVTYPE<- as.factor(nd4$EVTYPE)

nd4$EVTYPE <- factor(nd4$EVTYPE, levels = nd4$EVTYPE[order(-nd4$Cost)])  

g4<- ggplot(nd4,aes(EVTYPE,Cost),fill=Cost)+ geom_bar(stat="identity",color="black",fill="green")+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5))+ xlab("Event Type")+ ylab("Cost in Prop (Billion Dollars)")+ggtitle("Total Cost in Prop by Event Type (Billion)")+theme(plot.title = element_text(hjust = 0.5,size = 10, face = "bold"))


# Considering CROPDMG

nd5<- nd2 %>% group_by(EVTYPE) %>% summarise(Cost=sum(Cost)) %>% arrange(desc(Cost))
nd6<- slice(nd5,1:25)

# Transfor the cost column into "Billion dollars"

nd6$Cost<- nd6$Cost/1000000000
nd6$EVTYPE<- factor(nd6$EVTYPE,levels= nd6$EVTYPE[order(-nd6$Cost)])

g6<- ggplot(nd6,aes(EVTYPE,Cost),fill=Cost)+ geom_bar(stat="identity",color="black",fill="green")+ theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+xlab("Event Type")+ylab("Cost in Crop (Billion)")+ggtitle("Total Cost in Crop by Event Type (Billion Dollars)")+theme(plot.title = element_text(hjust = 0.5,size = 10, face = "bold"))


#Plot graph together

grid.arrange(g4,g6,ncol=2)

```

# Considering PROPDMG and CROPDMG in Together


```{r}
# Considering PROPDMG and CROPDMG together

nd7 <- rbind(nd4,nd6)
nd8 <- nd7 %>% group_by(EVTYPE) %>% summarise(Cost=sum(Cost)) %>% arrange(desc(Cost))
nd9 <- slice(nd8,1:25)

nd9$EVTYPE<- factor(nd9$EVTYPE,levels=nd9$EVTYPE[order(-nd9$Cost)])

g9 <- ggplot(nd9,aes(EVTYPE,Cost),fill=Cost)+ geom_bar(stat="identity",color="black",fill="green")+ theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+labs(title="Total Cost in Prop and Crop by Event Type (Billion)")+xlab("Event Type")+ylab("Cost in Prop and Crop (Billion Dollars)")+theme(plot.title = element_text(hjust = 0.5,size = 10, face = "bold"))

g9 

```


